---
title: "World Series Champions Simulation"
author: "Paul Forst, Ian O'Connor, Nathan Thomas"
date: "March 14, 2018"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r echo=FALSE, message=FALSE}
library(Lahman)
library(readr)
library(tidyverse)
source("game_sim.R")

options(stringsAsFactors = FALSE)
```

# The Rundown

This simulation collected statistics for the 16 Major League Baseball World Series champions from 2001 through 2016, created a game simulation and paired the teams up in games to see which team is the best.  The batting and pitching statistics are included for all teams from their respective championship seasons. 

To determine a true champion, the 16 teams will play a simulated season, playing all other teams twice, once at home and once away, in a round-robin format season. The simulation will run for 100 "seasons" and output a list of every season winner and their record in that particular season.

Probabilities for the simulation were based on modified "on-base" percentages which calculated the probability that any given batter would obtain one base, two bases, three bases, or four bases in an at bat or get out.  The top nine batters from each team were used to create a lineup.  These nine batters were selected by choosing the top 9 with the highest at bats through their respective regular season.  Pitching was factored into the simulation by calculating a similar "on-base" percentage, whereby producing an "apples-to-apples" environment for the simulation.  The on-base percentages were then adjusted by the average on-base percentage for pitchers on all 16 teams rosters.  This adjusted score was then weighted by the number of batters the pitcher faced throughout the season.  The scores were then aggregated creating a new base score for the entire pitching staff.  The new base scores were averaged and then the average was reduced from the base score creating a factor.  Negative factors meant that a pitching staff was very good and would have a negative effect on the hitters probabilities, while positive factors meant that a pitching staff was not as good and hitters would fare much better against the pitching staff.  

During the game simulation, the factor for the opposing pitching staff was multiplied against the hitting team's on-base probabilities.  The simulation for each game is then run.


```{r echo=FALSE, message=FALSE}
# Get information for the last X World Series champions
num_champs <- 16
Season <- c((max(Batting$yearID) - num_champs + 1):max(SeriesPost$yearID))
Winner <- NULL
for(i in Season){
      temp <- SeriesPost %>% 
            filter(yearID == i) %>% 
            filter(round == 'WS')
      Winner <- c(Winner, as.character(temp$teamIDwinner))
}

TeamId <- paste(Winner,Season,sep = "_")

#Remove factors from teamID and lgID field in Batting and Pitching data
Batting <- as.data.frame(Batting, stringsAsFactors = FALSE) %>% 
      filter(Batting$yearID >= min(Season)) %>% 
      mutate(teamID = as.character(teamID))
Pitching <- as.data.frame(Pitching, stringsAsFactors=FALSE) %>% 
      filter(Pitching$yearID >= min(Season)) %>% 
      mutate(teamID = as.character(teamID))

#WS Champs data frame
Champs <- data.frame(TeamId,Season,Winner, stringsAsFactors = FALSE)

####NOT NEEDED AT THIS TIME####
#Determine pitching OBP for each year
#yearly_obp <- Pitching %>% 
#      mutate(obp = ((H + BB + HBP + IBB + SH + SF)/(BFP))) %>% 
#      group_by(yearID)

#Pitching <- merge(Pitching, yearly_obp, by = "yearID")

battingStats <- NULL
pitchingStats <- NULL

# Loop to create the 16 teams
for (i in seq_along(Champs$TeamId)){
      temp <- Batting %>% 
            filter(yearID == Champs$Season[i]) %>% 
            filter(teamID == Champs$Winner[i]) %>% 
            arrange(-AB) %>% 
            mutate(TeamId = paste(teamID,yearID,sep = "_")) %>% 
            top_n(wt = AB, 9)
      
      battingStats[[i]] <- temp
      
      temp <- Pitching %>% 
            filter(yearID == Champs$Season[i]) %>% 
            filter(teamID == Champs$Winner[i]) %>% 
            arrange(-GS) %>% 
            mutate(obp = ((H + BB + HBP + IBB + SH + SF)/(BFP)), 
                   TeamId = paste(teamID,yearID,sep = "_"),
                   wt = ((BFP)/sum(BFP)))
      
      pitchingStats[[i]] <- temp
}

rm(Batting, Champs, Pitching, temp)

# Calculating Pitching Factor
LeaguePitchers <- NULL
for(i in 1:length(pitchingStats)){
      temp <- as.data.frame(pitchingStats[i])
      LeaguePitchers <- rbind(LeaguePitchers,temp)
}

AvePitch <- mean(LeaguePitchers$obp)
LeaguePitchers$LgAve <- AvePitch
LeaguePitchers$PitchFactor <- (LeaguePitchers$LgAve - LeaguePitchers$obp)/LeaguePitchers$LgAve
LeaguePitchers$WPF <- LeaguePitchers$PitchFactor*LeaguePitchers$wt
TeamPitch <- LeaguePitchers[,c(2,4,32,36)] %>% 
      group_by(yearID, teamID, TeamId) %>% 
      summarise(WPF = sum(WPF))

TeamPitch$WPFAve <- mean(TeamPitch$WPF) 
TeamPitch$WPFDiff <- TeamPitch$WPF - TeamPitch$WPFAve

rm(LeaguePitchers, temp)

#-----------------------------------------------
# Function to simulate home and home round robin
#-----------------------------------------------

#Set number of simulations to run
n <- 10
season_winners <- NULL

# Create progress bar
pb <- winProgressBar(title = "Running World Series Tournament: ", min = 0, max = n, width = 400)

for(sims in 1:n) {
      final_results <- NULL
      
      #Select team to be away and play all other teams
      for(i in 1:length(Season)){
            #Get away team and all opponents
            away <- TeamId[i]
            opponents <- TeamId[-i]
            
            #Gather away team stats
            away_batting <- as.data.frame(battingStats[i])
            away_pitching <- TeamPitch$WPFDiff[i]
            
            team_results <- NULL
            
            for(j in opponents) {
                  #Gather home team stats
                  index <- which(TeamId == j)
                  home_batting <- as.data.frame(battingStats[index])
                  home_pitching <- TeamPitch$WPFDiff[index]
                  
                  temp <- game_sim(away_batting, home_batting, away_pitching, home_pitching)
                  team_results <- rbind(team_results, temp)
                  
                  
            }
            
            final_results <- rbind(final_results, team_results)
            rm(away_batting, home_batting, away_pitching, home_pitching, team_results)
      }
      
      #Coerce scores to numeric and build standings
      final_results$away_score <- as.numeric(final_results$away_score)
      final_results$home_score <- as.numeric(final_results$home_score)
      final_results <- rbind(
            transmute(final_results,Team=home_team,Opponent=away_team,Home=TRUE,RunsFor=as.numeric(home_score),
                      RunsAgainst=away_score,Result=sign(RunsFor-RunsAgainst)),
            transmute(final_results,Team=away_team,Opponent=home_team,Home=FALSE,RunsFor=as.numeric(away_score),
                      RunsAgainst=home_score,Result=sign(RunsFor-RunsAgainst)))
      standings <- group_by(final_results,Team) %>% 
            summarize(Played=length(Team),
                      Won=sum(Result==1),
                      Lost=sum(Result==-1),
                      WinPct=(Won/Played),
                      Won10=sum(head(Result,10)==1),
                      Lost10=sum(head(Result,10)==-1),
                      HomeWon=sum(Result==1 & Home),
                      HomeLost=sum(Result==-1 & Home),
                      AwayWon=sum(Result==1 & !Home),
                      AwayLost=sum(Result==-1 & !Home),
                      RF=sum(RunsFor), RA=sum(RunsAgainst)
                      
            )
      finaltable <- transmute(standings,
                              Team,
                              Record = paste0(Won,"-",Lost),
                              HomeRec = paste0(HomeWon,"-",HomeLost),
                              AwayRec = paste0(AwayWon,"-",AwayLost),
                              WinPct = WinPct,
                              RunsFor=RF,
                              RunsAgainst=RA,
                              Last10 = paste0(Won10,"-",Lost10)) %>% 
            arrange(desc(WinPct))
      
      rm(final_results, standings)
      
      season_winners <- rbind(season_winners, finaltable[1,])
      
      # Update the progress bar
      setWinProgressBar(pb, sims, title=paste("Running World Series Tournament: ",round(sims/n*100, 1),"% done"))
}

rm(battingStats, pitchingStats)

close(pb)

season_winners
```

# Final Results

Once the simulation is finished running and we have each season's winner, we can aggregate these and output the overall champion, or the team that won the most simulations. 

```{r eval=TRUE, echo=FALSE}
champion <- season_winners %>% 
      group_by(Team) %>% 
      summarise(wins = n(),
                RunsScored = sum(RunsFor), 
                RunsAllowed = sum(RunsAgainst)) %>% 
      arrange(-wins, -RunsScored, RunsAllowed)

print(paste0("Your champion is ", champion[1,1]))
```
